A Mechanised Library for Generic (Co)Programming

### 2024-02-29 paper/paper.tex
minor
On basics about hylomorhisms.

\section{Algebras, Coalgebras and Hylomorphisms}
In the previous sections we have introduced the concept of a
\subsection{Algebras}

Algebras on container extensions are defined as expected. The carrier of the
algebra is \coq{A}, and the operations of the algebra are morphisms
of the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
Given a container extension \coq{F}, we define \emph{two}
(isomorphic) initial algebras. We will show the need for the second initial
algebra in the following section. Our fist initial algebra is defined as the
straightforward least fixpoint of the container extension \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words,
we would need to provide an equivalence for \coq{LFix}, but the
only way to do this is to define such equivalence simultaneously to the
definition for  \coq{LFix}. Such definition would be rejected by Coq.
There are other approaches that can circumvent this problem, by separating
the definition of least fixed points from the proofs that the functions that
get the contained elements are proper morphisms. But this leads to longer,
unnecessarily complicated proofs.  Furthermore, this extra layer interferes
with code extraction, in that it requires destructing/constructing subsequent
\coq{LFix} and  \coq{App}, and Coq may not always able to inline them.

>>>>

\section{Algebras, Coalgebras and Hylomorphisms}
Recursion schemes provide an abstract way to consume and generate data.  The
most simple and well-known recursion scheme is the \emph{fold} or
\emph{catamorphism} which consumes data. Its dual is the \emph{unfold} or
\emph{anamorphisms} which produces data.

For example, assuming the type of lists over a type \coq{A},
namely \coq{List A}, and given an
algebra \coq{alg : unit + A * List A -> List A} a
catamorphism \coq{cata alg : List A -> B} destructs a list into its head and
tail using a destructor \coq{InOp : List A -> unit + A * List A}. This latter is
actually an isomorphim with its inverse \coq{In : unit + A * List A -> List A}
defined by the maps \coq{Empty : unit -> List A}
and \coq{Cons : A * List A -> List A}.

After destructing the input the catamorphism is applying itsefl recursively.
This is done using functoriality of the shape of the data, which is a map of
type \coq{unit + A * List A -> unit + A * List B}. Once the recursion as
returned it result it applies the algebra \coq{alg} which composes together the
results of the recursion yielding the result of type \coq{B}.

More abstractly, assuming a base functor \coq{F} a catamorphism is the unique solution to the equation
\coq{cata alg = alg . fmap (cata alg) . InOp}.

% where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
% can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
% $F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
% $F$-coalgebra), $\InOp$ is the isomoprhism on the least fixed-point of $F$,
% namely $\InOp : \mu F \to F\mu F$, and $\OutOp$ is the isomorphism on the
% greatest fixed-point of $F$, namely $F \nu F \to \nu F$. When the unique
% solutions to the above equations exist we denote them as $\cata{a}$ and
% $\ana{c}$ respectively.

% C
Hylomorphism are more general recursion schemes capturing the
\emph{divide-and-conquer} idea. In short, the input is destructured by its not
structurally. In the example, we gave in the introduction the destructuring
program is given by \coq{divide} which is not an isomorphism.  The question the
hylomorphism satisfies is more general than the catamorphism in that the
destructing map \coq{InOp} is generalised to a \emph{recursive coalgebra}, that
is intuitively a coalgebra that can be applied only a finite amount of times.
In fact, in the case of \coq{divide} this is a recursive coalgebra because the
size of the list is getting smaller.

% This program is formed of essentially two parts. The first one is what is
% commonly thought as the \emph{partition} function destructuring the input list
% in three parts, the pivot and the other two sublists, and the second part
% recursing over the sublists and composing the results back together.

With this in mind we know proceed onto describing how to formalise hylomorphisms
in Coq.

\subsection{Algebras}
Algebras on container extensions are defined as expected. The carrier of the
algebra is \coq{A}, and the operations of the algebra are morphisms of the
following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
Given a container extension \coq{F}, we define \emph{two}
(isomorphic) initial algebras. We will show the need for the second initial
algebra in the following section. Our fist initial algebra is defined as the
straightforward least fixpoint of the container extension \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words,
we would need to provide an equivalence for \coq{LFix}, but the
only way to do this is to define such equivalence simultaneously to the
definition for  \coq{LFix}. Such definition would be rejected by Coq.
There are other approaches that can circumvent this problem, by separating
the definition of least fixed points from the proofs that the functions that
get the contained elements are proper morphisms. But this leads to longer,
unnecessarily complicated proofs.  Furthermore, this extra layer interferes
with code extraction, in that it requires destructing/constructing subsequent
\coq{LFix} and  \coq{App}, and Coq may not always able to inline them.
